---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Call out necessary packages

```{r}
    library(dplyr)
    library(ggplot2)
    library(lubridate)
```


### Function hist_and_stats 
#### Created for repeated generation of histogram and stats data with and without NAs 
```{r}
hist_and_stats <- function(my_csv){
    date_step <- with(my_csv, sapply(split(steps, date), sum, na.rm = TRUE))
    h <- hist(date_step, breaks = 5, plot = FALSE)
    h <- hist(date_step, breaks = 5, ylim = c(0, max(h$counts)), 
         main = "Steps Taken per Day", xlab = "Steps", xaxt = "n", yaxt ="n")
    axis(2, at = seq(0, max(h$counts), by = 2))
    axis(1, at = h$breaks)
    text(h$mids, h$counts, labels = h$counts, pos = 1)
    cat("\n\n#### What are mean and median total number of steps taken per day?\n")
    print(paste0("Mean = ", round(mean(date_step, na.rm = TRUE)), "; Median = ", median(date_step, na.rm = TRUE)))
}
```


### Loading and preprocessing the data
```{r}
    my_csv <- read.csv("activity.csv")
```

### Histogram of the total number of steps taken each day
```{r, results='asis'}
    hist_and_stats(my_csv)
```


### What is the average daily activity pattern?
```{r}
    five_min <- with(my_csv, sapply(split(steps, interval), mean, na.rm = TRUE))
    plot(as.numeric(names(five_min)), five_min, type = "l", main = "Average Steps per 5-min Interval", ylab = "Steps", xlab = "Interval")
    print(paste0("The 5-minute interval that, on average, contains the maximum number of steps = ", names(which.max(five_min))))
    
```


### Imputing missing values
Missing values are counted, and then imputed by the median of the specific interval.
```{r}
    print(paste0("Count of missing values = ",sum(is.na(my_csv))))
    my_csv <- my_csv %>% group_by(interval) %>% mutate(steps = ifelse(is.na(steps), median(steps, na.rm=TRUE), steps))
```

### Histogram of the total number of steps taken each day after missing values are imputed
```{r, results='asis'}
    hist_and_stats(my_csv)
```

The mean of the dataset changed, with the frequency and median remain unchanged.


#### Add weekofday factors and summarise
```{r}
    my_csv <- my_csv %>% mutate(weekofday = factor(ifelse(wday(date, week_start = 1) >= 6, "weekend", "weekday"), c("weekend","weekday")))
    wk_df <- my_csv %>%
        group_by(interval, weekofday) %>%
        summarise(steps = mean(steps), .groups = "drop")
```

#### Are there differences in activity patterns between weekdays and weekends?
```{r}
    ggplot(wk_df, aes(x = interval, y = steps)) +
        geom_line(color = "blue") +
        facet_wrap(~ weekofday, ncol = 1, strip.position = "top") +
        labs(x = "Interval", y = "Average Steps per Interval ") +
        theme(strip.background = element_rect(fill = "peachpuff"),
            panel.background = element_rect(fill = "white"))
```