## Title
#### Among a Total of 984 Events:
#### The Most Harmful Event is TORNADO, and the Greatest Economic Consequence Event is FLOOD

## Synopsis
This report analyzes storm event data from the **[NOAA Storm Database](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)**. Web interface is available **[here](https://www.ncei.noaa.gov/stormevents/)**.

We focus on:  
- Health impact: Fatalities and Injuries.   
- Economic impact: Property and Crop damages.  
  
Key steps:  
- Download and load the dataset. 
- Subset the dataset to relevant columns for efficiency.  
- Aggregate health and economic impacts by event type.  
- Visualize top contributors with bar charts and percentages.  
  
The following sections present detailed results with plots and commentary.  
  
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
  
#### Load required libraries (Messages loading the packages are suppressed on purpose)
```{r}
suppressMessages(library(dplyr))
suppressMessages(library(ggplot2))
suppressMessages(library(stringr))
```
  
#### Function 'report' is created for graphs and figures reporting wrt the scope
Refer to the **[Exponent Interpretation](https://github.com/shandrewlee/RepData/blob/master/PA2/exp%20Interpretation.md)** for justifications
```{r}
report <- function(agg_tbl, scope){
    top_items <- agg_tbl %>%
        slice_max(order_by = figure, n = 9)
    others <- agg_tbl %>%
        slice(-(1:9)) %>%
        summarise(EVTYPE = "OTHERS", figure = sum(figure)) 
    tbl <- bind_rows(top_items, others) %>%
        mutate(percentage = 100 * figure / sum(figure))
    
    print(paste0('Total number of events: ', length(agg_tbl$EVTYPE)))
    print(paste0('Events with Zero impact: ', sum(agg_tbl$figure == 0)))
    print(paste0('Events with impact: ', sum(agg_tbl$figure != 0)))
    
    print(paste0('The Event Impacting ',scope, ' the Most is ', top_items$EVTYPE[which.max(top_items$figure)]))
    
    ggplot(tbl, aes(x = reorder(EVTYPE, percentage), y = percentage)) +
        geom_bar(stat = "identity", fill = "cyan") +
        geom_text(aes(label = paste0(round(percentage, 1), "%")), hjust = 0.5, size = 3) +
        coord_flip() +
        labs(title = paste0("Top 10 Events Impacting ", scope), x = "Event Type", y = "Percentage") +
        theme_minimal(base_size = 10)
}
```
  
  
#### Function 'std_grammar' is create for standarding ambiguities and typos in EVTYPE
web_keyword is extracted from NOAA website for the 48 events in search list.<br>
custom_keyword is based on manual findings of ambiguities <br>
Irrelevant EVTYPE are dropped from the dataset<br>
```{r}
std_grammar <- function(data){
    
    data$EVTYPE <- toupper(data$EVTYPE)
    
    data$EVTYPE <- str_replace_all(data$EVTYPE, " {2,}", " ")
    
    #Correcting various typos for Thunderstorm Wind
    thunderstorm_typo <- 'T.*U.*E.*M WIN.*'
    data$EVTYPE[str_detect(toupper(data$EVTYPE), thunderstorm_typo)] <- 'THUNDERSTORM WIND'
    
    web_keyword <- c('ASTRONOMICAL LOW TIDE', 'AVALANCHE', 'BLIZZARD', 'COASTAL FLOOD', 'COLD/WINDHILL',
                     'DEBRIS FLOW', 'DENSE FOG', 'DENSE SMOKE', 'DROUGHT', 'DUST DEVIL', 'DUST STORM',
                     'EXCESSIVE HEAT', 'EXTREMEOLD/WINDHILL', 'FLASH FLOOD', 'FLOOD', 'FREEZING FOG', 
                     'FROST/FREEZE', 'FUNNEL CLOUD', 'HAIL', 'HEAT', 'HEAVY RAIN', 'HEAVY SNOW', 
                     'HIGH SURF', 'HIGH WIND', 'HURRICANE (TYPHOON)', 'ICE STORM', 'LAKE-EFFECT SNOW', 
                     'LAKESHORE FLOOD', 'LIGHTNING', 'MARINE HAIL', 'MARINE HIGH WIND', 'MARINE STRONG WIND', 
                     'MARINE THUNDERSTORM WIND', 'RIP CURRENT', 'SEICHE', 'SLEET', 'STORM SURGE/TIDE', 
                     'STRONG WIND', 'THUNDERSTORM WIND', 'TORNADO', 'TROPICAL DEPRESSION', 'TROPICAL STORM', 
                     'TSUNAMI', 'VOLCANIC ASH', 'WATERSPOUT', 'WILDFIRE', 'WINTER STORM', 'WINTER WEATHER')
    
    matches <- str_extract(toupper(data$EVTYPE), paste(web_keyword, collapse = "|"))
    data$EVTYPE <- ifelse(!is.na(matches), matches, data$EVTYPE)
    
    top_set <- data[data$EVTYPE %in% web_keyword,]  #This is the set cleaned by web_keyword
    bottom_set <- data[!data$EVTYPE %in% web_keyword,]  #This is the set to be cleaned by custom_keyword
    
    custom_keyword <- c('MICROBURST' = 'MICROBURST', 'AVALAN' = 'AVALANCHE',
                        'TSTM' = 'THUNDERSTORM WIND', 'THUNDERSTORM' = 'THUNDERSTORM WIND', 
                        'TYPHOON' = 'HURRICANE (TYPHOON)', 'HURRICANE' = 'HURRICANE (TYPHOON)',
                        'WET' = 'WET', 'RAIN' = 'RAIN', 'SNOW' = 'SNOW', 'SHOWER' = 'SHOWER', 
                        'MUD' = 'MUDSLIDE', 'LANDSLIDE' = 'LANDSLIDE', 'URBAN' = 'URBAN/SMALL STREAM', 
                        'WND' = 'WIND', 'COLD' = 'COLD','LOW TEMP' = 'COLD', 'HYPOTHERMIA' = 'COLD', 
                        'FREEZING' = 'FREEZING', 'HYPERTHERMIA' = 'HEAT', 'HIGH TEMP' = 'HEAT',
                        'WARM' = 'HEAT', 'HOT' = 'HEAT')
    
    matches <- str_extract((bottom_set$EVTYPE), paste(names(custom_keyword), collapse = "|"))
    bottom_set$EVTYPE <- ifelse(!is.na(matches), custom_keyword[matches], bottom_set$EVTYPE)
    
    #Dropping Irrelevant Event types
    bottom_set <- bottom_set[!grepl(".*county.*", bottom_set$EVTYPE, ignore.case = TRUE), ]
    bottom_set <- bottom_set[!grepl("^summary", bottom_set$EVTYPE, ignore.case = TRUE), ]
    bottom_set <- bottom_set[!grepl("^record", bottom_set$EVTYPE, ignore.case = TRUE), ]
    bottom_set <- bottom_set[!grepl("^monthly", bottom_set$EVTYPE, ignore.case = TRUE), ]
    bottom_set <- bottom_set[!bottom_set$EVTYPE %in% c('?','NONE'), ]
    
    # Combining the separate sets again
    data <- top_set
    data <- bind_rows(data, bottom_set)
}

```

### Data Processing
#### Download and Load the Source File
```{r, cache=TRUE}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
zip_file <- "stormdata.bz2"
if (!file.exists(zip_file)){
    download.file(url, zip_file, mode="wb")
    }
data <- read.csv(bzfile(zip_file))
```

### Extracting Effective Columns and Data Cleaning
```{r}
data <- data %>% 
    select(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)

data <- std_grammar(data)
```
  
#### Handling Exponent Values of PROPDMGEXP and CROPDMGEXP
```{r}
suppressWarnings(
data <- data %>%
    filter(!PROPDMGEXP %in% c('?','-')) %>%
    mutate(PROPDMG = case_when(
        PROPDMGEXP %in% c('0','1','2','3','4','5','6','7','8') ~ PROPDMG * 10 + as.numeric(PROPDMGEXP),
        PROPDMGEXP %in% c('H','h') ~ PROPDMG * 100,
        PROPDMGEXP %in% c('K','k') ~ PROPDMG * 1e+03,
        PROPDMGEXP %in% c('M','m') ~ PROPDMG * 1e+06,
        PROPDMGEXP %in% c('B','b') ~ PROPDMG * 1e+09,
        TRUE ~ PROPDMG))
)

suppressWarnings(
data <- data %>%
    filter(!CROPDMGEXP %in% c('?','-')) %>%
    mutate(CROPDMG = case_when(
        CROPDMGEXP %in% c('0','1','2','3','4','5','6','7','8') ~ CROPDMG * 10 + as.numeric(CROPDMGEXP),
        CROPDMGEXP %in% c('H','h') ~ CROPDMG * 100,
        CROPDMGEXP %in% c('K','k') ~ CROPDMG * 1e+03,
        CROPDMGEXP %in% c('M','m') ~ CROPDMG * 1e+06,
        CROPDMGEXP %in% c('B','b') ~ CROPDMG * 1e+09,
        TRUE ~ CROPDMG))
)
```
  
### Aggregate health and economic impacts by event type
```{r}
evt_health <- data %>% group_by(EVTYPE) %>% 
    summarise(figure = sum(FATALITIES + INJURIES, na.rm = TRUE), .groups = "drop")
evt_econ <- data %>% group_by(EVTYPE) %>% 
    summarise(figure = sum(PROPDMG + CROPDMG, na.rm = TRUE), .groups = "drop")
```
  
  
## Results  
### Bar Charts for Health and Economy Impact
```{r}
report(evt_health, 'Health (FATALITIES + INJURIES)')
report(evt_econ, 'Economy (PROPDMG + CROPDMG)')
```
