## Title
#### Among a Total of 984 Events:
#### The Most Harmful Event is TORNADO, and the Greatest Economic Consequence Event is FLOOD

## Synopsis
This report analyzes storm event data from the **[NOAA Storm Database]("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2")**. Web interface is available **[here](https://www.ncei.noaa.gov/stormevents/)**.

We focus on:  
- Health impact: Fatalities and Injuries.   
- Economic impact: Property and Crop damages.  
  
Key steps:  
- Download and load the dataset. 
- Subset the dataset to relevant columns for efficiency.  
- Aggregate health and economic impacts by event type.  
- Visualize top contributors with bar charts and percentages.  
  
The following sections present detailed results with plots and commentary.  
  
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
  
#### Load required libraries (Messages loading the packages are suppressed on purpose)
```{r}
suppressMessages(library(dplyr))
suppressMessages(library(ggplot2))
```
  
#### Function 'report' is created for graphs and figures reporting wrt the scope
```{r}
report <- function(agg_tbl, scope){
    top_items <- agg_tbl %>%
        slice_max(order_by = figure, n = 9)
    others <- agg_tbl %>%
        slice(-(1:9)) %>%
        summarise(EVTYPE = "OTHERS", figure = sum(figure)) 
    tbl <- bind_rows(top_items, others) %>%
        mutate(percentage = 100 * figure / sum(figure))
    
    print(paste0('Total number of events: ', length(agg_tbl$EVTYPE)))
    print(paste0('Events with Zero impact: ', sum(agg_tbl$figure == 0)))
    print(paste0('Events with impact: ', sum(agg_tbl$figure != 0)))
    
    print(paste0('The Event Impacting ',scope, ' the Most is ', top_items$EVTYPE[which.max(top_items$figure)]))
    
    ggplot(tbl, aes(x = reorder(EVTYPE, percentage), y = percentage)) +
        geom_bar(stat = "identity", fill = "cyan") +
        geom_text(aes(label = paste0(round(percentage, 1), "%")), hjust = 0.5, size = 3) +
        coord_flip() +
        labs(title = paste0("Top 10 Events Impacting ", scope), x = "Event Type", y = "Percentage") +
        theme_minimal(base_size = 10)
}
```
  
### Data Processing
#### Download and Load the Source File
```{r, cache=TRUE}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
zip_file <- "stormdata.bz2"
if (!file.exists(zip_file)){
    download.file(url, zip_file, mode="wb")
    }
data <- read.csv(bzfile(zip_file))

data <- data %>% 
    select(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
```
  
#### Handling Exponent Values of PROPDMGEXP and CROPDMGEXP
```{r}
suppressWarnings(
data <- data %>%
    filter(!PROPDMGEXP %in% c('?','-')) %>%
    mutate(PROPDMG = case_when(
        PROPDMGEXP %in% c('0','1','2','3','4','5','6','7','8') ~ PROPDMG * 10 + as.numeric(PROPDMGEXP),
        PROPDMGEXP %in% c('H','h') ~ PROPDMG * 100,
        PROPDMGEXP %in% c('K','k') ~ PROPDMG * 1e+03,
        PROPDMGEXP %in% c('M','m') ~ PROPDMG * 1e+06,
        PROPDMGEXP %in% c('B','b') ~ PROPDMG * 1e+09,
        TRUE ~ PROPDMG))
)

suppressWarnings(
data <- data %>%
    filter(!CROPDMGEXP %in% c('?','-')) %>%
    mutate(CROPDMG = case_when(
        CROPDMGEXP %in% c('0','1','2','3','4','5','6','7','8') ~ CROPDMG * 10 + as.numeric(CROPDMGEXP),
        CROPDMGEXP %in% c('H','h') ~ CROPDMG * 100,
        CROPDMGEXP %in% c('K','k') ~ CROPDMG * 1e+03,
        CROPDMGEXP %in% c('M','m') ~ CROPDMG * 1e+06,
        CROPDMGEXP %in% c('B','b') ~ CROPDMG * 1e+09,
        TRUE ~ CROPDMG))
)
```
  
### Aggregate health and economic impacts by event type
```{r}
evt_health <- data %>% group_by(EVTYPE) %>% 
    summarise(figure = sum(FATALITIES + INJURIES, na.rm = TRUE), .groups = "drop")
evt_econ <- data %>% group_by(EVTYPE) %>% 
    summarise(figure = sum(PROPDMG + CROPDMG, na.rm = TRUE), .groups = "drop")
```
  
  
## Results  
### Bar Charts for Health and Economy Impact
```{r}
report(evt_health, 'Health')
report(evt_econ, 'Economy')
```
